---
title: "MRIP pure recreational catch distribution"
author: "Marc Nadon"
execute: 
  warning: false
  error: false
format: 
 html:
  embed-resources: true
editor: visual
page-layout: article
---

This script is to obtain the distribution of Deep7 and total per-trip catches for
non-commercial fishers. To do so, we acquired the MRIP catch interviews from 2003
to 2022 where the question "Do you ever sell your catch?" was recorded. This allow
us to filter catch interviews conducted by pure recreational fishers and explore
the distribution of their catch.

## Preliminary steps

### Load packages and define root directory

```{r}
suppressWarnings( require(pacman,quietly=T) )
pacman::p_load("tidyverse","data.table","this.path","readxl")
root_dir <- here(..=1)

# Quiet the summarize function messages, 
# add to 16 digits to load the interview_id correctly.
options(dplyr.summarise.inform = FALSE,digits=16)
```

### Load the necessary data sets

We obtained the MRIP interviews from Hongguang Ma.

```{r}
IT <-fread(
        fs::path(root_dir,"01_Data","MRIP never sell interviews.csv")) %>% 
         select(interview_id=ID_CODE,sp_mrip_id=SP_CODE,num_caught=catch) %>% 
           mutate(sp_mrip_id=as.character(sp_mrip_id))

SP <- fread(fs::path(root_dir,"01_Data","mrip_mean_weights.csv")) %>% 
           mutate(sp_mrip_id=as.character(sp_mrip_id)) %>% 
              arrange() %>% as.data.table()
```

Let's merge the species information

```{r}
IT <- IT %>% left_join(SP,by="sp_mrip_id") 
```

Let's split up the information contained in the *interview_id*.

```{r}
# Remove empty spaces.
IT$interview_id <- str_replace_all(IT$interview_id," ","")

# Split the date out.
IT <- IT %>% 
       mutate(interview_date=
        lubridate::ymd(str_sub(interview_id,start=6,end=nchar(interview_id)-3)),
              year=lubridate::year(interview_date))

IT <- IT %>% mutate(lbs_caught=num_caught*mrip_mean_lbs) %>% 
           select(interview_id,year,sci_name,sp_frs_id,
           sp_mrip_id,num_caught,lbs_caught)
```

Let's see what the distributions look like globally.

```{r}
# Catch quantiles, per species and positive catch only.
Qt.sp <- IT %>% group_by(sp_frs_id) %>%
  reframe(quant=round(quantile(lbs_caught,c(0.90,0.95,0.99,1)),0),
                      quantile=c("q90","q95","q99","max")) %>%
  pivot_wider(id_cols=,names_from=quantile,values_from=quant)

# Catch quantiles for D7 catch (positive-only)
IT.D7 <- IT %>% filter(sp_frs_id!="s20") %>% group_by(interview_id, year) %>%
            summarize(num_caught=sum(num_caught),lbs_caught=sum(lbs_caught))

Qt.D7 <- IT.D7 %>% group_by() %>% 
  reframe(quant=round(quantile(lbs_caught,c(0.90,0.95,0.99,1)),0),
          quantile=c("q90","q95","q99","max")) %>% 
              pivot_wider(names_from=quantile,values_from=quant)

# Put everything back together.
Qt.D7$sp_frs_id <- "d7"

Final <- rbind(Qt.sp,Qt.D7) %>% left_join(SP,by="sp_frs_id") %>%
  select(sci_name,sp_frs_id,q90,q95,q99,max) %>% as.data.table()

Final[is.na(sci_name)]$sci_name <- "Deep7"

write_csv(Final,fs::path(root_dir,"03_Outputs","mrip_quantiles.csv"))
```

Some plots.

```{r}
# Remove rare species.
ITS <- IT %>% filter(!(sp_frs_id %in% c("s20","s58","s15")))
FS  <- Final %>% filter(!(sp_frs_id %in% c("s58","s15")))

# Per-species in numbers
ggplot(data=ITS,aes(x=num_caught))+
  geom_histogram(binwidth=1)+
  #geom_vline(data=FS[sci_name!="Deep7"],aes(xintercept=Q99),col="red")+
  facet_wrap(~sci_name,scales="free")+
  theme_minimal(base_size=16)+
  labs(x="Number caught per trip",y="Number of trips")

ggsave(last_plot(),file=fs::path(root_dir,"03_Outputs","MRIP_Sp_Num_Caught.png"),
       height=4,width=8,dpi=300)
    
# Per-spcies in pounds
ggplot(data=IT,aes(x=lbs_caught))+geom_histogram(binwidth=2)+
  facet_wrap(~sci_name,scales="free")
  
# D7 totals
ggplot(data=IT.D7,aes(x=num_caught)) + geom_histogram(binwidth=1)
ggplot(data=IT.D7,aes(x=lbs_caught)) + geom_histogram(binwidth=5) +
  geom_vline(data=FS[sci_name=="Deep7"],aes(xintercept=q99),col="red")+
  theme_bw(base_size=14)+
  labs(x="Deep7 catch per trip (lbs)",y="Number of trips")

ggsave(last_plot(),file=fs::path(root_dir,"03_Outputs","MRIP_D7_Lbs_Caught.png"),
       height=4,width=8)
   
```