---
title: "Lamson 2007 pure recreational catch distribution"
author: "Marc Nadon"
format: html
editor: visual
---

This script is to obtain the distribution of Deep7 and total per-trip catches for non-commercial fishers. To do so, we acquired the MRIP catch interviews from 2003 to 2022 where the question "Do you ever sell your catch?" was recorded. This allow us to filter catch interviews conducted by pure recreational fishers and explore the distribution of their catch.

## Preliminary steps

### Load packages and define root directory

```{r}
suppressWarnings( require(pacman,quietly=T) )
pacman::p_load("tidyverse","data.table","this.path","readxl")

root_dir <- here(..=1)

# Quiet the summarize function messages.
options(dplyr.summarise.inform = FALSE)
```

### Load the necessary data sets

Explanations for the columns in the survey file can be found in the "Lamson - BF Printable Codes Final II.docx" file. In short, SurveyID represents unique fishers answering the questionnaire. Q21 is "do you ever sell bottomfish?" (allows us to classify CML and non-CML fishers), with "1" equal yes and "2" equal no. The rest of the columns selected below are yearly catch in pounds per Deep 7 species and uku.

```{r}
LM <-fread(
        fs::path(root_dir,"01_Data","Lamson - BF Survey Table FINAL.csv")) %>% 
         select(fisher_id=SurveyID,
                sell_bf=Q21,
                s21=Q62, s97=Q66, s15=Q70, s17=Q74,
                s58=Q78, s22=Q82, s19=Q86, s20=Q90) %>% as.data.table()

SP <- fread( fs::path(root_dir,"01_Data","mrip_mean_weights.csv")) %>% 
         select(sci_name,sp_frs_id)
```

Let's convert code "99" and "no" to NA and make all columns numeric.

```{r}
LM <- LM %>%  
  mutate(across(where(is.character), ~ if_else(. %in% c("no","99"),NA,.))) %>% 
  mutate(across(where(is.numeric), ~ if_else(. %in% 99,NA,.))) %>% 
  mutate(across(everything(), ~ as.numeric(.)))
```

Let's calculate the bottomfish (+uku) and Deep7 totals. Let's also remove all fishers not catching any of these.

```{r}
LM <- LM %>% mutate(d7=rowSums(across(c(s21:s19)),na.rm=T),
                    allBF=rowSums(across(c(s21:s20)),na.rm=T))

LM <- LM %>% filter(!is.na(allBF)&allBF!=0)
```

Let's pivot species-specific catch longer.

```{r}
LM <- LM %>% 
  pivot_longer(cols=(s21:allBF),names_to="sp_frs_id",values_to="lbs_caught") %>% 
   as.data.table()
```

Now let's calculate the quantile of these annual catches, for non-CML fishers.

```{r}
LM <- LM %>% left_join(SP,by="sp_frs_id") %>% 
  mutate(sci_name=if_else(sp_frs_id=="allBF","allBF",sci_name),
         sci_name=if_else(sp_frs_id=="d7","Deep7",sci_name))

# Filter for "never sell" and remove zero annual catches.
LM <- LM %>% filter(sell_bf==2,lbs_caught>0) 

Qt.sp <- LM %>% group_by(sci_name,sp_frs_id) %>%
  reframe(quant=round(quantile(lbs_caught,c(0.90,0.95,0.99,1)),0),
          quantile=c("q90","q95","q99","max")) %>% 
  pivot_wider(id_cols=,names_from=quantile,values_from=quant) 

Qt.sp <- Qt.sp %>% select(sci_name,sp_frs_id,q90:max)

write_csv(Qt.sp,file.path(root_dir,"03_Outputs","lamson_quantiles.csv"))
```

Let's create some plots.

```{r}
# Remove rare species and uku.
LMS <- LM %>% filter(!(sp_frs_id %in% c("allBF","s58","s15")))
QS <- Qt.sp %>% filter(!(sp_frs_id %in% c("allBF","s58","s15")))

# Per-species in pounds
ggplot(data=LMS,aes(x=lbs_caught))+
  geom_histogram()+
  geom_vline(data=QS,aes(xintercept=q90),col="red")+
  geom_vline(data=QS,aes(xintercept=q95),col="darkorange")+
  geom_vline(data=QS,aes(xintercept=q99),col="darkgreen")+
  geom_text(data=QS,aes(label=q90,x=q90,y=Inf),hjust=1.2,vjust=2)+
  geom_text(data=QS,aes(label=q95,x=q95,y=Inf,),hjust=0.8,vjust=2)+
  geom_text(data=QS,aes(label=q99,x=q99,y=Inf),hjust=0.5,vjust=2)+
  facet_wrap(~sci_name,scales="free")+
  theme_bw(base_size=16)+
  labs(x="Annual pounds caught",y="Number of fishers")

ggsave(last_plot(),file=fs::path(root_dir,"03_Outputs","Lamson_Sp_Lbs_Caught.png"),
       height=8,width=10,dpi=300)
```
