---
title: "Annual Deep 7 catch from the Bottomfish Vessel Registry"
author: "Marc Nadon"
format: html
---

## General description

This script creates annual distributions of species-specific mean-catch-per-vessel
by county using a selection of fishers from the Fishing Reporting System (FRS) to
serve as non-commercial proxies. This involves classifying all cml_no x year
records as either "commercial" or "non-commercial", based on a series of filters :
a) cml owner conducted a trip with a high catch in a given year, with per-trip max
catch trigger points derived from MRIP "never-sell" interviews (see script
03_MRIP_interviews_catch.qmd), and b) cml owner had a high catch per year, derived
from Lamson (2007) mail surveys. These filters are applied at both the species
level and all D7 together.

The script does this by 1) loading a list of bottomfish fishers and establishing
the number of active fishers in a given year, 2) loading catch reports contained
in the Fisher Reporting System (FRS), 3) selecting a subset of fishers from the
FRS (i.e. non-commercial fisher proxies) , and 4) sampling annual catches by
county from these fishers. The end product are distributions of non-commercial
Deep 7 catches in Hawaii, by year.

### Load packages and define root directory

```{r}
#| warning: false
require(pacman,quietly=T); pacman::p_load("tidyverse","data.table","this.path",
                              "readxl","flextable","ggridges","ggthemes")
root_dir <- here(..=1)
dir.create(root_dir,"03_Outputs")
set_flextable_defaults(na_str="N/A",table.layout="autofit",line_spacing=0.5)
options(dplyr.summarise.inform = FALSE)
```

## Enter the limits that define non-commercial proxies in the FRS

These limits are derived from the MRIP fishing trip interviews (2004-2022) and
represent the 99th percentile of catches, when the catches are not zero.

```{r}
# Per-trip limits (MRIP interviews 2004-2022)
d7_lbs_per_trip     <- 94
uku_lbs_per_trip    <- 50
paka_lbs_per_trip   <- 70
onaga_lbs_per_trip  <- 108
ehu_lbs_per_trip    <- 40
kale_lbs_per_trip   <- 28
gindai_lbs_per_trip <- 21
  
# Per-year limits (Lamson 2007, mail surveys, "never sell BF")
d7_lbs_per_year     <- 486
uku_lbs_per_year    <- 300
paka_lbs_per_year   <- 213
onaga_lbs_per_year  <- 150
ehu_lbs_per_year    <- 180
kale_lbs_per_year   <- 83
gindai_lbs_per_year <- 82
```

### Load the necessary data sets

We need the registered fisher list and the FRS data.

```{r}
# Load fisher list from 01_Registered_fisher_list.qmd
FL <- read_xlsx(fs::path(root_dir,"03_Outputs","Hawaii Registered Fishers.xlsx"))

# Load fisher counts from 01_Registered_fisher_list.qmd.
FC <- fread(fs::path(root_dir,"03_Outputs","Fisher_counts.csv")) %>% 
              filter(cml_registr=="N")

# Load FRS data.
FS <- read_rds(fs::path(root_dir,"01_Data","HFY1994_2023.rds")) %>%
         mutate(year=year(FISHED_DATE)) %>% filter(year>=2011) %>% 
         as.data.table()
```

### Clean up the FRS catch data and merge with fisher information

Select FRS columns and rename certain variables.

```{r}
F1 <- FS %>%  select(cml_no.fs=FISHER_LIC_FK,
                     year,
                     fished_date=FISHED_DATE,
                     area_fk=AREA_FK,
                     gear_fk=GEAR_FK,
                     hours=HOURS_FISHED,
                     num_gear=NUM_GEAR,
                     landing_port_fk=LANDING_PORT_FK,
                     fisher_fname=FISHER_FNAME,
                     fisher_lname=FISHER_LNAME,
                     species_fk=SPECIES_FK,
                     num_kept=NUM_KEPT,
                     lbs_kept=LBS_KEPT,
                     lbs_sold=LBS_SOLD)

# Only select Deep-sea handline trip (>98% of D7 are caught on this gear)
F1 <- F1 %>% filter(gear_fk==3)

F1[species_fk==36]$species_fk <- 21 #Set ehu code 36 to new code 21
F1$species_fk <- paste0("s",F1$species_fk)
```

Now, let's join catch data with fisher information from the Hawaii Registered
Fisher list.

```{r}
F1 <- F1 %>% left_join(FL,by=join_by(cml_no.fs==cml_no,year))

cat("A small number of records didn't match: ",nrow(unique(F1[is.na(county)])))

F1 <- F1 %>% filter(!is.na(county))
```

[NOTE: there are about 50 cml_no with no county info. Checking transaction list,
these are CML that didn't have a vessel_id listed... and said Captain==N, but
still reported D7 ("no vessel" cmls were filtered early on in Step 1, thus the
lack of match). What to do with these?]{.underline}

Create unique identifiers for trips and sum catch by species

```{r}
F1 <- F1 %>% mutate(trip_id=paste0(cml_no.fs,"_",fished_date)) %>% 
  group_by(year,cml_no.fs,cml_registr,bf_registr,county,trip_id,species_fk) %>%
       summarize(lbs_kept=sum(lbs_kept))
```

Calculate the aggregated catch per trip

```{r}
# Pivot the trip catch in lbs by species.
F2 <- F1 %>%
  pivot_wider(names_from=species_fk,values_from=lbs_kept,values_fill=0) %>%
      as.data.table()

# Deep7 catch
F2$d7 <- F2$s15+F2$s17+F2$s19+F2$s21+F2$s22+F2$s58+F2$s97

# All species
F2$allsp <- rowSums(F2[,7:ncol(F2)])

# Drop all other species columns except D7
F2 <- F2 %>% select(year:trip_id,s20,s15,s17,s19,s21,s22,s58,s97,d7,allsp)
```

Save this Fisher trip information for the Shinny app used to explore the different
options available to generate non-commercial catch.

```{r}
write_csv(F2,fs::path(root_dir,"03_Outputs","FRS_trips.csv"))
```

Classify all CMLs as either "commercial" (C) or "non-commercial" (NC), based on
the per-trip limits defined previously.

```{r}
# All trips classified as non-commercial as a starting point.
F2$trip_type <- "nc_trip"

# This classification will change to commercial if catches go higher than one
# of these trigger points.
F2[d7  > d7_lbs_per_trip|
   s17 > kale_lbs_per_trip| 
   s19 > paka_lbs_per_trip| 
   s21 > ehu_lbs_per_trip|
   s22 > onaga_lbs_per_trip|
   s97 > gindai_lbs_per_trip| 
   s20 > uku_lbs_per_trip 
   ]$trip_type <- "comm_trip"

# Create a separate data set to see trip stats by county and annual catch
F2a <- F2 %>% select(year,cml_no.fs,county,trip_id,trip_type) %>% 
             group_by(year,cml_no.fs,county,trip_type) %>%
             summarize(n=n()) %>%
             pivot_wider(names_from=trip_type,values_from=n,values_fill=0) %>% 
             as.data.table()

# Classify CML x year combinations into fisher types.
F2a$fisher_type               <- "Comm"
F2a[comm_trip==0]$fisher_type <- "NC"

# Summary
F2b <- F2a %>% group_by(year) %>% 
         summarize(nc_trip=sum(nc_trip),comm_trip=sum(comm_trip)) 

ggplot(data=F2b,aes(x=year))+geom_line(aes(y=nc_trip),col="blue")+
  geom_line(aes(y=comm_trip),col="red")+ylab("# of trips")

ggsave(last_plot(),file=fs::path(root_dir,"03_Outputs","Trips_all.png"),
       height=4,width=8)
```

Quick check on the distribution of trip catches for all BF + caught a D7 fishers.

```{r}
F2 %>% filter(d7!=0&bf_registr=="Y"&year==2023) %>% ggplot(aes(x=d7))+
          geom_histogram(breaks=c(seq(0,500,by=10),max(F2$d7)),
                         position="identity")+
          labs(x="Deep7 catch per trip (lbs)",y="Number of trips")+
          coord_cartesian(xlim=c(0,480))+theme_bw(base_size = 15)

ggsave(last_plot(),file=fs::path(root_dir,"03_Outputs","Comm_Trip_2024_D7.png"),
       height=5,width=8)
```

Let's sum the trip catch by *year* and *cml_no*, and merge the fisher information
from the step above with this data set.

```{r}
F3 <- F2 %>% group_by(year,cml_no.fs,bf_registr,cml_registr,county) %>% 
             summarize_if(is.numeric,sum)

F3 <- F3 %>% left_join(F2a,join_by(cml_no.fs,year,county)) %>%
              relocate(nc_trip:fisher_type,.after=cml_no.fs) %>% as.data.table()
```

Now we can apply the annual catch filters derived from Lamson (2007).

```{r}
# This classification will change to commercial if catches go higher than one of these trigger points.

#cat("Number of NC fishers, pre-filter: ",nrow(F3[fisher_type=="NC"]),"\n")

#F3[d7  > d7_lbs_per_year|
#   s17 > kale_lbs_per_year| 
#   s19 > paka_lbs_per_year| 
#   s21 > ehu_lbs_per_year|
#   s22 > onaga_lbs_per_year|
#   s97 > gindai_lbs_per_year| 
#   s20 > uku_lbs_per_year 
#   ]$fisher_type <- "comm_trip"

#cat("Number of NC fishers, post-filter: ",nrow(F3[fisher_type=="NC"]))
```

An important step is to decide which fishers to remove from this data set. One
filter we discussed was to remove fishers that did not catch a single D7 during a
year. Another discussion is to just keep fishers registered on the BFVR.

```{r}
cat("Prop. caught D7: ",round(nrow(F3[d7>0])/nrow(F3),2),"\n")
cat("Prop. caught D7 and BF registered: ",
             round(nrow(F3[d7>0&bf_registr=="Y"])/nrow(F3),2))
```

Important step: decide which filters to apply. Let's start with keep fishers D7\>0
and BF registered only.

```{r}
F3 <- F3 %>% filter(d7!=0&bf_registr=="Y")

#F3 <- F3 %>% filter(d7!=0)
```

NOTE: the difference between these two filters appear to be about 29,000 lb vs
30,000 lb yearly in average catch. Not as big of a deal as originally thought.

Now let's only keep the "non-commercial" fisher proxies and clean up the data.

```{r}
F4 <- F3 %>% filter(fisher_type=="NC") %>% 
              select(year,cml_no.fs,county,6:allsp) %>% 
                arrange(year,county) 
```

## Generate annual non-commercial catch distribution

Now that we have 1) the number of active non-commercial BF fisher by year, and 2)
a proxy data set of annual catch per "non-commercial" fisher, we can sample "n"
times from that annual catch data set, where "n" is the number of non-comm. BF
fishers in a specific *year* and *county*. This will generate a final distribution
of non-commercial catches by year.

```{r}
# Add the # of NC vessels by year x County to the catch data
F4 <- F4 %>% left_join(FC,by=join_by(year,county)) %>% 
                relocate(n_bf_fishers,.after=county)

# Check the number of fishers to sample from in each county x year with the number of active NC D7 fishers from Step 1.
num.frs  <- as.data.table(table(F4$year,F4$county)) %>% 
            rename(year=V1,county=V2,n_frs=N) %>% mutate(year=as.numeric(year))
chck.num <- D3 %>% inner_join(num.frs,by=join_by(year,county))

write_csv(chck.num,file=fs::path(root_dir,"03_Outputs","Frs_vs_BFVR.csv"))

# Resample the catch data by the number of vessels in a year X county block. Since there are often less fishers to sample than the n target, we need to sample with replacement. 
Results <- list()
for(i in 1:1000){
 aSample <- F4 %>% group_by(year,county) %>% 
                    sample_n(n_bf_fishers[1],replace=T) %>% 
                      add_column(iter=i) %>% relocate(iter,.before=year)
 Results <- append(Results,list(aSample))
}

Results <- rbindlist(Results)
```

Sum fisher-specific annual catches by iteration.

```{r}
Final <- Results %>% group_by(iter,year,county) %>%
           summarize_at(vars(s20:allsp),sum)
```

Plot results.

```{r}
# Simple line plot
Final %>% group_by(year,county) %>% summarize_at(vars(s20:allsp),mean) %>% 
    ggplot(aes(x=year,y=d7))+
      geom_line()+geom_point()+
      facet_wrap(~county)+theme_bw(base_size=16)+
      theme(axis.text.x=element_text(angle=45,hjust=1))+
      labs(x="Year",y="Deep7 non-commercial catch (lbs)")

suppressMessages(ggsave(last_plot(),
  file=fs::path(root_dir,"03_Outputs","NC_county_simple.png"),width=8,height=6))

# Box plot
ggplot(data=Final,aes(x=factor(year),y=d7))+
  geom_boxplot(fill="lightblue",alpha=0.6,outlier.size=0.6)+
  facet_wrap(~county)+theme_minimal()+
  theme(axis.text.x=element_text(angle=45,hjust=1))+
  labs(x="Year",y="Deep7 non-commercial catch (lbs)")

# Ridge plot
suppressMessages(print( 
      ggplot(data=Final,aes(y=factor(year),x=d7))+
      geom_density_ridges(scale=1,fill="lightblue",
                      quantile_lines=T,quantiles=2,
                      rel_min_height=0.0001)+
      theme_minimal(base_size=13)+
      theme(axis.text.x=element_text(angle=45,hjust=1),
            strip.text = element_text(size = 14))+
      coord_flip()+labs(y="Year",x="Deep7 non-commercial catch (lbs)")+
      facet_wrap(~county,ncol=2)
))

suppressMessages(ggsave(last_plot(),
      file=fs::path(root_dir,"03_Outputs","NC_county.png"),width=8,height=6))
```

Finally, sum county-level catches by year and generate catch tables.

```{r}
Final.b <- Final %>% group_by(iter,year) %>% summarize_at(vars(s20:allsp),sum)
Final.c <- Final.b %>% group_by(year) %>% summarize_at(vars(s20:allsp),mean)

# Some per-species summary
Final.b.sp <- Final.b %>% pivot_longer(
                cols=s20:s97,names_to="species",values_to="lbs_caught") %>%
                  filter(species %in% c("s19","s21","s22")) %>%
                     select(iter,year,species,lbs_caught)

# Some extra summary for the Deep7 group
Final.d <- Final.b %>% group_by(year) %>%
                  summarize(d7.meanlog=mean(log(d7)),
                            d7.sdlog=sd(log(d7)),
                            d7.mean=mean(d7))

write_csv(Final.d,file=fs::path(root_dir,"03_Outputs","NC_BF_all.csv"))
```

Some figures and comparison with the CML catch.

```{r}
# Species figure
suppressMessages(print( 
      ggplot(data=Final.b.sp,aes(y=factor(year),x=lbs_caught))+
      geom_density_ridges(scale=1,fill="lightblue",
                      quantile_lines=T,quantiles=2,
                      rel_min_height=0.0001)+
      theme_minimal(base_size=13)+
      theme(axis.text.x=element_text(angle=45,hjust=1),
            strip.text = element_text(size = 14))+
      coord_flip()+labs(y="Year",x="Non-commercial catch (lbs)")+
      facet_wrap(~species,ncol=3)
      ))

# Deep 7 figure
suppressMessages(print(ggplot()+
      geom_density_ridges(data=Final.b,aes(y=factor(year),x=d7),
                          scale=1,fill="lightblue",
                          quantile_lines=T,quantiles=2,
                          rel_min_height=0.0001)+
      theme_minimal(base_size=14)+
      theme(axis.text.x=element_text(angle=45,hjust=1))+
      coord_flip()+labs(y="Year",x="Deep7 non-commercial catch (lbs)")
))
suppressMessages(ggsave(last_plot(),width=8,height=5,
              file=fs::path(root_dir,"03_Outputs","NC_BF_all.png")))

# Simple Deep 7 figure
# Simple line plot
Final.b %>% group_by(year) %>% summarize_at(vars(s20:allsp),mean) %>% 
    ggplot(aes(x=year,y=d7))+
      geom_line()+geom_point()+
      theme_bw(base_size=16)+ylim(0,50000)+
      scale_x_continuous(breaks=seq(2011,2023,by=1))+
      theme(axis.text.x=element_text(angle=45,hjust=1))+
      labs(x="Year",y="Deep7 non-commercial catch (lbs)")

suppressMessages(ggsave(last_plot(),
  file=fs::path(root_dir,"03_Outputs","NC_simple.png"),width=8,height=6))
```